---
title: "Protein-level preprocessing, imputation, and supplementary QC"
output: html_document
---

## Summary

This document records the protein-level preprocessing, quality control, and
imputation workflow used for the manuscript, including the
run-level PCA in supplementary figure 1. 
Input quantities are already aggregated to protein groups by DIA-NN; no precursor -> protein aggregation is performed when handling the protein group matrix (pg_matrix)

## Data inputs

Run this document from the repository root after configuring `data_dir` and
`output_dir` below. Required inputs are:

- `full_analysis2/report.pg_matrix.tsv`: DIA-NN output (in PRIDE), the protein-group intensity matrix
- `Input/mapnamesreport.csv`: has `original_names` and `new_name` columns, which map the experimental layout (plate-well) to subject-fraction (plate-well-subject-fraction) so that technical and individual-level variables can be combined
- `Input/contaminants.csv`: list of contaminants to exclude with `Accession` and `Gene_symbol`.
- At the end, for the supplementary PCA, a `demogs` dataset containing `Subject`, `Group`(CON/RES/AD/DEM), `age_death`, `pmi`, `study`, and `msex`. It is external to this analysis*
  
*Subject-level inputs and generated objects are not distributed with the public
code. ROSMAP participant-level clinical and neuropathological data can be requested through the Rush Alzheimer’s Disease Center at https://www.radc.rush.edu and AD Knowledge Portal at https://www.synapse.org, subject to the applicable data-use agreements. 

## Output

`MAI_prot_model_new.rds` is a SummarizedExperiment storing both the unimputed log2-scaled assay,
`log2Quant` and the imputed log2-scaled assay, `MAI_imputed`, alongside respective protein and run-level metadata. Both assays are 7,271 proteins x 524 runs after QC processing

Filtering order, exclusions, transformation, random seed, imputation settings,
and PCA settings are retained from the original workflow. The original
session-dependent imputation references are made explicit: `log2_mat` supplies
`log2Quant`, and `mat` is then extracted from that assay in `se_log2`. This
connection should be checked against the original analysis object.

## Setup and protein matrix import

```{r import-and-contaminants}
library(tidyverse)
library(data.table)
library(readr)

#configure local paths (data dir contains input, output dir is local output)
data_dir <- "data" # <- define input path here 
output_dir <- "outputs_local"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# -------------------------------
# 1) Read wide protein matrix
# - data.table, chosen for wrangling
# -------------------------------
f <- file.path(data_dir, "report.pg_matrix.tsv")
prot_wide <- as.data.table(read_tsv(f))

#identify metadata columns
meta_cols <- c(
  "Protein.Group",
  "Protein.Names",
  "Genes",
  "First.Protein.Description")

# all others columns are LC-MSMS runs
run_cols <- setdiff(names(prot_wide), meta_cols)

# -------------------------------
# 2) Pivot wide -> long
# -------------------------------
prot <- melt(
  prot_wide,
  id.vars = meta_cols,
  measure.vars = run_cols,
  variable.name = "Run",
  value.name = "Quant",
  variable.factor = FALSE)

rm(prot_wide) 

# -------------------------------
# 3) Clean run names
# -------------------------------
prefix <- "^.*/" # Strip the acquisition-directory prefix from each run path
prot[, Run := sub(prefix, "", Run)]
prot[, Run := sub("\\.mzML$", "", Run)]
prot[, Run := sub("_[0-9]{14}(?=\\s|$)", "", Run, perl = TRUE)]
prot[, Run := gsub(" ", "_", Run)]

#extract plate/well experimental identity
tmp <- sub(".*P([0-9]+)_W([0-9]+).*", "\\1;\\2", prot$Run)
spl <- tstrsplit(tmp, ";")
prot[, `:=`(
  Plate_num = as.integer(spl[[1]]),
  Well_num  = as.integer(spl[[2]])
)]
rm(tmp, spl)

prot[, RunID := paste0("P", Plate_num, "_W", Well_num)]

#exclude well 17 (not relevant to this experiment)
prot <- prot[RunID != "P7_W17"]

# -------------------------------
# 4) Map experimental layout (plate-well) to subject-fraction
# -------------------------------
mapnames <- fread(file.path(data_dir, "Input", "mapnamesreport.csv"))

setnames(mapnames, "original_names", "RunID_old")
setnames(mapnames, "new_name", "RunID_new")

setkey(prot, RunID)
setkey(mapnames, RunID_old)

prot[mapnames, RunID := fcoalesce(i.RunID_new, RunID)]

# split mapped RunID
prot[, c("Plate","Well","Subject","Fraction") :=
       tstrsplit(RunID, "-", fixed = TRUE, fill = NA)]

#remove trailing 'p' from fraction
prot[, Fraction := sub("p$", "", Fraction)]

# -------------------------------
# 5) Remove contaminants
# -------------------------------
contaminants <- fread(file.path(data_dir, "contaminants.csv")) #read in from data directory

# counts BEFORE
n_prot_before <- uniqueN(prot$Protein.Group)
n_gene_before <- uniqueN(prot$Genes)

prot <- prot[
  !grepl("cRAP", Protein.Group) & #flagged as being in contaminant .fasta file in DIANN search
    endsWith(Protein.Names, "_HUMAN") & 
    !(Protein.Group %in% contaminants$Accession | 
        Genes %in% contaminants$Gene_symbol)
]

# counts AFTER
n_prot_after <- uniqueN(prot$Protein.Group)
n_gene_after <- uniqueN(prot$Genes)

print("unique proteins before vs after contaminant removal:")
n_prot_before; n_prot_after
print("unique Genes before vs after contaminant removal:")
n_gene_before; n_gene_after

# drop helper columns
prot[, c("Run","Plate_num","Well_num") := NULL]
```
## Sample exclusions and protein filtering

```{r sample-and-protein-filtering}
library(data.table)

# -------------------------------
# 6) Remove technical samples
# -------------------------------
prot <- prot[
  !is.na(Fraction) &
    Fraction != "Pool" &
    !Subject %in% c("A8","C5","E3","G1","J8") &
    RunID != "P4-W76-G2-2p" #failed run excluded (>80% missingness)
]

# Counts after removal of technical samples
n_prot_after <- uniqueN(prot$Protein.Group)
n_gene_after <- uniqueN(prot$Genes)

# -------------------------------
# 7) Remove proteins with ≥70% missingness
# -------------------------------
protein_missingness <- prot[, .(
  prop_missing = mean(is.na(Quant))
), by = Protein.Group]

keep_proteins <- protein_missingness[prop_missing < 0.7, Protein.Group]

prot <- prot[Protein.Group %in% keep_proteins]

# counts after filtering
print("Proteins vs Genes after removal of entries with >=70% missingness")
uniqueN(prot$Protein.Group)
uniqueN(prot$Genes)

# -------------------------------
# 8) Remove entries listing multiple protein accessions or gene symbols 
# - these are non-proteotypic entries
# -------------------------------

prot <- prot[
  !grepl(";", Protein.Group, fixed = TRUE) & # Exclude semicolon-separated protein accessions
    !grepl(";", Genes, fixed = TRUE) 
]
print("Proteins vs Genes after removal of non-proteotypic entries")
uniqueN(prot$Protein.Group)
uniqueN(prot$Genes)
```
## Missingness after filtering

```{r missingness-summary}
prot[, .(mean_missing = mean(is.na(Quant)))] #~0.15%
```

## Construct the protein matrix
```{r construct-long-matrix}
# -------------------------------
# 9) Protein-level matrix
# one row per protein × run
# -------------------------------

prot_report <- prot[, .(
  Quant = if (all(is.na(Quant))) NA_real_ else Quant), by = .(
  Protein.Group,
  First.Protein.Description,
  Genes,
  RunID,
  Plate,
  Well,
  Subject,
  Fraction)]

setcolorder(prot_report,  c("RunID","Plate","Well","Subject","Fraction", "Protein.Group","First.Protein.Description","Genes","Quant"))

#set row order (subject A1-J8, fraction 1-7)
prot_report[, `:=`(
  row1 = substr(Subject, 1, 1), #split Subject into row letter and numeric column
  row2 = as.integer(substring(Subject, 2))
)]

prot_report[, `:=`(
  row1 = factor(row1, levels = LETTERS[1:10]), #letters 1-10 are A–J
  row2 = as.integer(row2),
  Fraction = factor(Fraction, levels = as.character(1:7))
)]

setorder(prot_report, row1, row2, Fraction, Protein.Group)

# ------------------------------------------------------------
# create a run order table (one row per run)
# ------------------------------------------------------------
run_order <- unique(prot_report[, .(RunID, Plate, Well, Subject, Fraction, row1, row2)])
setorder(run_order, row1, row2, Fraction)
prot_report[, c("row1", "row2") := NULL] #drop helper columns afterwards

# ------------------------------------------------------------
# 10) cast to wide protein x run matrix
# one row per Protein.Group
# ------------------------------------------------------------
# make sure prot_report is a data.table
setDT(prot_report)

assay_dt <- dcast(prot_report,
  Protein.Group + First.Protein.Description + Genes ~ RunID,
  value.var = "Quant") #back to wide format

#keep metadata separate
rowdata <- unique(assay_dt[, .(Protein.Group, First.Protein.Description, Genes)])

#numeric assay matrix
assay_mat <- as.matrix(assay_dt[, -(1:3), with = FALSE])
rownames(assay_mat) <- assay_dt$Protein.Group #set row names as protein group
storage.mode(assay_mat) <- "double" #store as a double object

# reorder columns to desired A1 1-7 ... J8 1-7 order
desired_runs <- run_order$RunID
assay_mat <- assay_mat[, desired_runs, drop = FALSE]

rm(prot_report, assay_dt)
```

## Log2 transformation and MAI imputation
```{r transform-and-impute}
library(data.table)
library(SummarizedExperiment)
library(S4Vectors)
library(MAI)

# ------------------------------------------------------------
# 11) log2 transform
# ------------------------------------------------------------
assay_mat[assay_mat <= 0] <- NA_real_ ##assumes Quant is positive intensity, any zeros -> NA
log2_mat <- log2(assay_mat)

# ------------------------------------------------------------
# 12) build colData (run-level) and rowData (protein-level) metadata for summarizedExperiment (SE)
# ------------------------------------------------------------
coldata <- as.data.frame(run_order[, .(RunID, Plate, Well, Subject, Fraction)])
rownames(coldata) <- coldata$RunID

rowdata_df <- as.data.frame(rowdata)
rownames(rowdata_df) <- rowdata_df$Protein.Group

#align rowdata to assay
rowdata_df <- rowdata_df[rownames(log2_mat), , drop = FALSE]

# ------------------------------------------------------------
# 13) create unimputed SE
# ------------------------------------------------------------
se_log2 <- SummarizedExperiment(
  assays = SimpleList(log2Quant = log2_mat),
  colData = DataFrame(coldata),
  rowData = DataFrame(rowdata_df))

se_log2 #524 runs (after failed-run exclusion) x 7271 proteins
```
## Mechanism aware imputation
- MAI is run with `MCAR_algorithm = "BPCA"`, `MNAR_algorithm = "Single"`,
300 random-forest trees, `proximity = FALSE`, and seed 137690. Both
unimputed and imputed log2-scaled matrices are retained in the saved SE.

```{r}
# ------------------------------------------------------------
# 14) MAI imputation (mechanism-aware imputation)
# ------------------------------------------------------------
#extract matrix from SE
mat <- assay(se_log2, "log2Quant") #from hereon, use the unimputed assay extracted from the SE

rm(log2_mat) #original log-scaled matrix can be removed

set.seed(137690)

log2se_0.7_MAI <- MAI( 
  data_miss = mat,
  MCAR_algorithm = "BPCA", #MCAR algorithm to use (for predicted random missingness)
  MNAR_algorithm = "Single", #MNAR algorithm to use (for predicted non-random missingness)
  assay_ix = 1, #if SE, designates the assay to impute
  forest_list_args = list(ntree = 300, proximity = FALSE), verbose = TRUE) #random forest arguments for training

imputed <- log2se_0.7_MAI$Imputed_data

# restore dimnames explicitly
dimnames(imputed) <- dimnames(mat)

# add imputed assay alongside unimputed assay
assay(se_log2, "MAI_imputed") <- imputed
se_log2
```


## Save the processed analysis object

```{r save-object}
# ——————————————————— Now save —————————————————————————
#saveRDS(se_log2, file = file.path(output_dir, "MAI_prot_model_new.rds"))
```
## Supplementary run-level PCA

This chunk produces the supplementary S1 PCA figure. Datapoints are coloured by fraction,
plate, pre-imputation missingness, sex, age at death, and post-mortem interval.
This requires the external restricted `demogs` dataset specified at the beginning*

```{r supplementary-pca}
library(SummarizedExperiment)
library(dplyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(viridis)

# ============================================================
# 1. Load SummarizedExperiment
# ============================================================

se_log2_07 <- readRDS(file.path(output_dir, "MAI_prot_model_new.rds"))
assayNames(se_log2_07)

imputed_assay   <- "MAI_imputed"
unimputed_assay <- "log2Quant"

# ============================================================
# 2. Extract log2 matrices
# ============================================================

mat_imputed <- assay(se_log2_07, imputed_assay)
mat_unimputed <- assay(se_log2_07, unimputed_assay)

stopifnot(
  identical(dim(mat_imputed), dim(mat_unimputed)),
  identical(colnames(mat_imputed), colnames(mat_unimputed)))

#confirm that the imputed matrix has no missing values
sum(is.na(mat_imputed))

#proportion of proteins missing in each run before imputation
prop_na <- colMeans(is.na(mat_unimputed))

# ============================================================
# 3. Remove proteins unsuitable for PCA
# ============================================================

# Retain proteins with finite values across all runs
keep_complete <- rowSums(!is.finite(mat_imputed)) == 0

# Remove proteins with zero variance across runs
protein_var <- apply(mat_imputed, 1, var, na.rm = TRUE)
keep_variable <- is.finite(protein_var) & protein_var > 0

mat_pca <- mat_imputed[keep_complete & keep_variable, , drop = FALSE]

dim(mat_pca)

# ============================================================
# 4. Run PCA
# ============================================================

pca <- prcomp(
  t(mat_pca), #transpose
  center = TRUE, #centre
  scale. = FALSE)

variance_explained <- 100 * pca$sdev^2 / sum(pca$sdev^2)

pc1_var <- round(variance_explained[1], 1)
pc2_var <- round(variance_explained[2], 1)

pc1_var
pc2_var

# ============================================================
# 5. Assemble sample-level metadata
# ============================================================

run_pca_df <- as.data.frame(pca$x[, 1:2, drop = FALSE]) %>%
  rownames_to_column("SampleID") %>%
  left_join(
    colData(se_log2_07) %>%
      as.data.frame() %>%
      rownames_to_column("SampleID"),
    by = "SampleID") %>%
  mutate(
    Prop_NA = prop_na[SampleID]) %>%
  left_join(
    demogs %>%
      dplyr::select(
        Subject,
        Group,
        age_death,
        pmi,
        study,
        msex) %>%
      distinct(Subject, .keep_all = TRUE),
    by = "Subject",
    suffix = c("", ".demogs"))

# ============================================================
# 6. Format plotting variables
# ============================================================

run_pca_df <- run_pca_df %>%
  mutate(
    Plate = factor(
      Plate,
      levels = paste0("P", 1:7)),

    Fraction_plot = factor(
      as.character(Fraction),
      levels = as.character(1:7)),

    Sex_plot = case_when(
      msex == 0 ~ "0",
      msex == 1 ~ "1",
      TRUE ~ "NA"),

    Sex_plot = factor(
      Sex_plot,
      levels = c("0", "1", "NA")))

# ============================================================
# 7. Shared plotting settings
# ============================================================

x_lab <- paste0("PC1 (", pc1_var, "%)")
y_lab <- paste0("PC2 (", pc2_var, "%)")

base_theme <- theme_bw(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(
      linewidth = 0.25,
      colour = "grey88" ),
    plot.title = element_text(
      size = 10,
      face = "plain",
      hjust = 0),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.position = c(0.98, 0.03),
    legend.justification = c(1, 0),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.7),
      colour = "grey60"))

p_fraction <- ggplot(
  run_pca_df,
  aes(PC1, PC2, colour = Fraction_plot)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(
    title = "Fraction",
    x = x_lab,
    y = y_lab,
    colour = "Fraction") +
  base_theme +
  theme(
    legend.position = c(0.98, 0.03),
    legend.direction = "horizontal")


p_plate <- ggplot(
  run_pca_df,
  aes(PC1, PC2, colour = Plate)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(title = "Plate",
    x = x_lab,
    y = y_lab,
    colour = "Plate") +
  scale_colour_viridis_d(
    option = "H",
    na.value = "grey70") +
  guides(colour = guide_legend(
      nrow = 2)) +
  base_theme +
  theme(
    legend.position = c(0.98, 0.03),
    legend.direction = "horizontal")


p_missing <- ggplot(
  run_pca_df,
  aes(PC1, PC2, colour = Prop_NA)) +
  geom_point(size = 2.2, alpha = 0.9) +
  labs(
    title = "Missingness",
    x = x_lab,
    y = y_lab,
    colour = "Prop. NA") +
  scale_colour_viridis_c(
    option = "magma",
    direction = 1,
    na.value = "grey70") +
  guides(
    colour = guide_colourbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(2.3, "cm"),
      barheight = unit(0.35, "cm"))) +
  base_theme +
  theme(legend.direction = "horizontal")


p_sex <- ggplot(
  run_pca_df,
  aes(PC1, PC2, colour = Sex_plot)) +
  geom_point(size = 2.2, alpha = 0.85) +
  labs(
    title = "Sex",
    x = x_lab,
    y = y_lab,
    colour = "Sex") +
  scale_colour_manual(
    values = c(
      "0" = "#F8766D",
      "1" = "#00BFC4",
      "NA" = "grey65"),
    drop = FALSE) +
  guides(colour = guide_legend(
      nrow = 1,
      override.aes = list(size = 3, alpha = 1))) +
  base_theme +
  theme(legend.direction = "horizontal")


p_age <- ggplot(
  run_pca_df,
  aes(PC1, PC2, colour = age_death)) +
  geom_point(size = 2.2, alpha = 0.9) +
  labs(
    title = "Age at death",
    x = x_lab,
    y = y_lab,
    colour = "Age") +
  scale_colour_viridis_c(
    option = "G",
    na.value = "grey70") +
  guides(colour = guide_colourbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(2.3, "cm"),
      barheight = unit(0.35, "cm"))) +
  base_theme +
  theme(legend.direction = "horizontal")


p_pmi <- ggplot(
  run_pca_df,
  aes(PC1, PC2, colour = pmi)) +
  geom_point(size = 2.2, alpha = 0.9) +
  labs(title = "Post-mortem interval",
    x = x_lab,
    y = y_lab,
    colour = "PMI") +
  scale_colour_viridis_c(
    option = "rocket",
    direction = -1,
    na.value = "grey70") +
  guides(
    colour = guide_colourbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(2.3, "cm"),
      barheight = unit(0.35, "cm"))) +
  base_theme +
  theme(
    legend.direction = "horizontal")

# ============================================================
# 8. Combine panels
# ============================================================

pca_figure <- (p_fraction + p_plate) / 
  (p_missing + p_sex) / 
  (p_age + p_pmi) +
  plot_annotation(
    title = "PCA of run-level protein intensities",
    subtitle = "log2(intensities), MAI imputed",
    theme = theme(
      plot.title = element_text(
        size = 10,
        face = "bold"),
      plot.subtitle = element_text(size = 10)))

pca_figure
```



